Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

### 2020-11-10 doc_arXiv/doc_arXiv.tex
revised doc shared with co-authors

\author{
    Byron C. Jaeger
    \thanks{Source code available at
\url{https://github.com/bcjaeger/INTERMACS-missing-data}}
   \\
    Department of Biostatistics \\
    University of Alabama at Birmingham \\
  Birmingham, AL 35211 \\
  \texttt{\href{mailto:bcjaeger@uab.edu}{\nolinkurl{bcjaeger@uab.edu}}} \\
  }

@@

\begin{abstract}
\textbf{Background} Risk predictions play an important role in clinical
decision making. When developing risk prediction models, practitioners
often impute missing values to the mean. The purpose of this article is
to evaluate the impact of applying different strategies to impute
missing values on the prognostic accuracy of prediction models fitted to
the imputed data. A secondary objective was to compare the accuracy of
different imputation methods. To complete these objectives, we used data
from the Interagency Registry for Mechanically Assisted Circulatory
Support (INTERMACS). \newline\textbf{Methods and Results} We applied
thirteen different strategies to impute missing values in combination
with three different strategies to fit a risk prediction model for
mortality and transplant after receiving mechanical circulatory support.
Model performance was measured by 12-month discrimination, calibration
and net reclassification index. Results indicated that multiple
imputation consistently provided prediction models with greater
prognostic accuracy than other missing data strategies, particularly
imputation to the mean. \newline\textbf{Conclusion} Selecting an optimal
strategy to handle missing values impacts the prognostic accuracy of
downstream models. In the current analysis, multiple imputation emerged
as an optimal strategy to handle missing values in the INTERMACS data.
The current study shows that evaluation and selection of an optimal
strategy to impute missing data has the potential to improve prognostic
accuracy of risk predictions for other longitudinal registries.
\end{abstract}

\keywords{
    Missing Data,
   \and
    INTERMACS,
   \and
    imputation,
   \and
    heart failure,
   \and
    mortality,
   \and
    risk prediction
  }

@@

Heart disease is a leading cause of death in the United States. Heart
failure, a primary component of heart disease, affects over 6 million
Americans, and for \textasciitilde10\% of these patients medical
management is no longer effective
\cite{benjamin2017heart,national2017health}. Mechanical circulatory
support (MCS) is a surgical intervention in which a mechanical device is
implanted in parallel to the heart to improve circulation
\cite{patel2014contemporary}. Typically, MCS is used while a patient
waits for a heart transplant (bridge-to-transplant) or in some cases as
an alternative to transplant (destination therapy)
\cite{slaughter2009advanced}. Over 250,000 patients could benefit from
MCS \cite{miller2011left}. However, less than 4,000 new patients receive
a long-term MCS device each year, with widely heterogeneous outcomes
\cite{stewart2011keeping}. The 2-year survival on MCS ranges from 61\%
for destination therapy to 78\% for bridge-to-transplant
\cite{patel2014contemporary}. Therefore, there is great need for
reliable predictions of patient-specific risk to experience adverse
events after receiving MCS. This information can be used to improve
patient selection for MCS, inform the design of next generation pumps,
and refine patient management strategies.

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{kirklin2017eighth, kormos2019society, Adamo950}. A critical step
for developing risk prediction models is handling missing data, which is
common in a ``real world'' database that is not designed for the rigor
of data completeness found in a clinical trial. While some missing
values are related to data entry error, others may relate to instability
of a patient's circulatory status at the time of implant. For instance,
hemodynamic laboratory values are missing for patients in whom invasive
catheterizations were not performed. Patients without invasive
catheterizations may be too ill (unstable) to tolerate the procedure.
Instead, such patients may, after determination of basic hemodynamics,
proceed directly to device placement.

@@

In this section, we describe the INTERMACS Registry in Section
\ref{subsec:intermacs} and provide a broad overview of general
objectives for imputation of missing data in the context of statistical
inference and supervised statistical learning in Section
\ref{subsec:inference_and_learning}. We then describe thirteen
strategies to handle missing values (Section \ref{subsec:imputation})
and three strategies to develop risk prediction models after missing
data have been imputed (Section \ref{subsec:modeling}). Next, we
describe how risk prediction models and by extension the imputation
strategy applied before training the model were evaluated in Section
\ref{subsec:evaluation}. We outline an internal validation procedure
that was applied to compare combinations of imputation and modeling
strategies using hierarchical Bayesian models in Section
\ref{subsec:internal}. Last, details on computation and software
packages used for the current analysis are described in Section
\ref{subsec:computing}

@@

The INTERMACS data is publicly available on biolincc at
\url{https://biolincc.nhlbi.nih.gov/studies/intermacs/}. INTERMACS is a
North American observational registry for patients receiving MCS devices
that began as a partnership between the National Heart, Lung, and Blood
Institute, US Food and Drug Administration, the Centers for Medicaid and
Medicare Services, industry, and individual hospitals with the mission
of improving MCS outcomes. In 2018, INTERMACS became an official Society
of Thoracic Surgeons database.

The current analysis was conducted using publicly available data
provided by the National Heart, Lung, and Blood Institute. We included a
contemporary cohort of 14,738 patients who received continuous flow LVAD
from 2012-2017. Patient follow-up begins after implantation of a
durable, long term MCS device and continues while the device is in
place. Registry endpoints include death on a device, heart
transplantation, or cessation of support (for recovery and non-recovery
reasons). INTERMACS collects pre-implant patient characteristics,
medical status, laboratory values, and many other variables. Data is
collected at regularly scheduled follow-up as well as during adverse
events such as re-hospitalization. This is a secondary analysis of
de-identified data obtained from the National Heart Lung and Blood
Institute. Primary data collection is approved through University of
Alabama Institutional Review Board and at individual sites.

@@

Previous research has thoroughly investigated statistical inference in
the context of missing data. In this setting, analysts often create
multiple imputed datasets, replicate an analysis in each of them, and
then pool their results to obtain valid test statistics for hypothesis
testing \cite{rubin2004multiple}. Imputation strategies that create a
single imputed dataset (\textit{e.g., }imputation to the mean) have been
shown to increase type I errors (\textit{i.e., }rejecting a true null
hypothesis) for inferential statistics by artificially reducing the
variance of observed data and ignoring the uncertainty attributed to
missing values \cite{van2018flexible}. To ensure valid inference,
imputation models should leverage outcome variables to predict missing
predictors \cite{sterne2009multiple}. When very few data are missing,
analysts may apply listwise deletion, \textit{i.e., }removing any
observation with at least one missing value. However, listwise deletion
can easily lead to biased inference \cite{van2020rebutting}.

@@

Several of the imputation methods we considered allowed for creation of
one or many imputed datasets. For downstream models fitted to multiple
sets of imputed data, we applied the modeling technique to each imputed
training set, separately, and then used a pooling technique to generate
a single set of predictions based on multiple imputed testing data.
Specifically, we created 10 imputed training and testing sets, then
applied a modeling procedure to each imputed training set and computed
model predictions on the corresponding imputed testing set, which led to
10 sets of predictions. To aggregate these predictions, we computed the
median for each observation. Informal experiments where the mean was
used instead of the median to aggregate predictions showed little or no
difference in the model's prediction accuracy.

@@

Imputing data to the mean involves three steps. First, numeric and
nominal variables are identified. Second, for each numeric variable, the
mean is computed and used to impute missing values in the corresponding
variable. Third, for each nominal variable, the mode is computed
(because one cannot compute a mean for categorical variables) and used
to impute missing values in the corresponding variable. The computed
means and modes are then stored for future imputation of testing data.
Because imputation to the mean is frequently used in practice, we use it
as a reference for comparison of all other strategies to impute missing
values,

@@

Imputation with Bayesian regression draws imputed values from the
posterior distribution of model parameters, accounting for uncertainty
in both model error and estimation \cite{rubin2004multiple}. Multiple
imputation with chained equations (MICE), a well known technique for
generating multiply imputed data
\cite{azur2011multiple, van2018flexible, van2006fully}. In each
iteration of MICE, each specified variable in the dataset is imputed
using the other variables in the dataset. These iterations are run until
convergence criteria have been met.

@@

\hypertarget{evaluating-imputation-accuracy}{%
\subsection{Evaluating imputation
accuracy}\label{evaluating-imputation-accuracy}}

@@

Imputation accuracy was computed for each numerical and nominal
variable, separately. Numeric variable imputation accuracy was measured
using a re-scaled mean-squared error:
\[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is smaller than
\(\textrm{MSE}(\textrm{imputation to the mean})\), equal to 0 if the two
MSEs are equal, and less than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is greater than
\(\textrm{MSE}(\textrm{imputation to the mean})\). Nominal variable
imputation accuracy was measured using a re-scaled classification
accuracy:
\[1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if the classification error of the current
imputation method is less than that of imputation to the mean, equal to
0 if the two classification errors are equal, and less than 0 of the
classification error of the current imputation method is worse than
imputation to the mean. These numeric and nominal scores are analogous
to the more well known \(R^2\) and Kappa statistics, respectively, but
are modified slightly in the current analysis so that imputation to the
mean will always have a score of 0. This modification makes it easier to
compare the accuracy of each imputation strategy directly with
imputation to the mean.

@@

\hypertarget{computational-details}{%
\subsection{Computational details}\label{computational-details}}

@@

Our source code is available on GitHub (see
\url{https://github.com/bcjaeger/INTERMACS-missing-data}).

@@

Adjusting for the amount of additional missing data amputed and the
outcome variable, the posterior probability that an imputation strategy
would improve the scaled Brier score of a downstream model relative to
imputation to the mean was maximized by using multiple imputation with
random forests (probability of improvement: 0.998; Figure
\ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
random forests was estimated to have the highest posterior probability
of improving the C-index in comparison to using imputation to the mean
(probability of improvement: 0.962; Figure
\ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
probability of a reduction in calibration was 1 when using either single
or multiple imputation with random forests compared with imputation to
the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}). Although
imputation using random forest was estimated to be the best overall
option, there was moderate to strong evidence that MIA and each multiple
imputation strategy we applied would improve prognostic accuracy of
downstream models compared with imputation to the mean.

@@

In this article, we leveraged data from the INTERMACS registry to
evaluate how the use of different imputation strategies prior to fitting
a risk prediction model would impact the external prognostic accuracy of
the model. External prognostic accuracy was measured at 6 months after
receiving MCS to focus on short term risk prediction, and the primary
measure of accuracy was the scaled Brier score. We evaluated the
performance of 12 imputation strategies in a broad range of settings by
varying the type of modeling algorithm, the outcome variable, and the
amount of additional missing data amputed prior to performing
imputation. Our resampling experiment indicated that conducting multiple
imputation has a high likelihood of increasing the downstream scaled
Brier score and C-index of risk prediction models compared with
imputation to the mean. Additionally, multiple imputation with random
forests emerged as the imputation strategy that maximized the
probability of developing a more prognostic model compared with
imputation to the mean.

In previous studies involving the INTERMACS data registry, imputation to
the mean has been applied prior to developing a mortality risk
prediction model
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society}.
The current analysis has shown that the prognostic accuracy of these
models can be increased by implementing a different strategy to impute
missing values, particularly if multiple imputation is applied. However,
few studies have investigated best practices on the use of multiple
imputation for prediction models, and few software packages allow users
to implement multiple imputation procedures that adhere to best
practices for prediction model development (\textit{e.g., }using only
the training data to impute missing values). Further development of open
source software that allows investigators to safely and succinctly
implement multiple imputation for prediction models (\textit{e.g., }the
\texttt{miceRanger} package) may increase the presence of multiple
imputation in published risk prediction models.

\clearpage
\begin{table}
\caption{Participant characteristics stratified by event status.}
\label{tbl_characteristics}

\begin{tabular}{lccccc}
\toprule
Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
\midrule
No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
months\_post\_implant & 18 (16) & 25 (18) & 16 (11) & 13 (14) & 13 (10)\\
Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
Sex &  &  &  &  & \\
\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
\addlinespace
\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
Race &  &  &  &  & \\
\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
\addlinespace
Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
Device strategy &  &  &  &  & \\
\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
\addlinespace
LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
Urinary albumin, g/dl & 3.41 (0.64) & 3.42 (0.61) & 3.41 (0.64) & 3.33 (0.66) & 3.49 (0.65)\\
Urinary creatinine, mg/dl & 1.39 (0.68) & 1.36 (0.62) & 1.17 (0.53) & 1.49 (0.77) & 1.35 (0.67)\\
CV pressure & 10.8 (6.2) & 10.4 (6.1) & 9.9 (6.0) & 11.6 (6.5) & 10.7 (6.0)\\
Periphal edema & 35 & 33 & 37 & 44 & 31\\
\addlinespace
BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
Bilirubin levels, mg/dl & 1.37 (1.84) & 1.25 (1.14) & 1.28 (1.95) & 1.52 (2.46) & 1.41 (1.97)\\
Device type &  &  &  &  & \\
\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
\addlinespace
Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
CPB time, minutes & 95 (49) & 93 (45) & 88 (38) & 103 (56) & 90 (44)\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
\end{tabular}

>>>>

\author{
    Byron C. Jaeger
    \thanks{Source code available at
\url{https://github.com/bcjaeger/INTERMACS-missing-data}}
   \\
    Department of Biostatistics \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{\href{mailto:bcjaeger@uab.edu}{\nolinkurl{bcjaeger@uab.edu}}} \\
   \And
    Ryan Cantor, PhD
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Venkata Sthanam, MS
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Rongbing Xie, DrPH
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    James K. Kirklin, MD
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{} \\
   \And
    Ramaraju Rudraraju, PhD
   \\
    University of Alabama at Birmingham \\
  Birmingham, AL \\
  \texttt{\href{mailto:ramaraju@uab.edu}{\nolinkurl{ramaraju@uab.edu}}} \\
  }

@@

\begin{abstract}
\textbf{Background} Risk prediction models play an important role in
clinical decision making. When developing risk prediction models,
practitioners often impute missing values to the mean. We evaluated the
impact of applying other strategies to impute missing values on the
prognostic accuracy of downstream risk prediction models, i.e., models
fitted to the imputed data. A secondary objective was to compare the
accuracy of imputation methods based on artificially induced missing
values. To complete these objectives, we used data from the Interagency
Registry for Mechanically Assisted Circulatory Support (INTERMACS).
\newline\textbf{Methods and Results} We applied twelve imputation
strategies in combination with two different modeling strategies for
mortality and transplant risk prediction following surgery to receive
mechanical circulatory support. Model performance was evaluated using
Monte-Carlo cross validation and measured based on outcomes 6-months
following surgery using the scaled Brier score, concordance index, and
calibration error. We used Bayesian hierarchical models to compare model
performance. Multiple imputation with random forests emerged as a robust
strategy to impute missing values, increasing model concordance by 0.003
(25th, 75th percentile: 0.008, 0.052) compared with imputation to the
mean for mortality risk prediction using a downstream proportional
hazards model. The posterior probability that single and multiple
imputation using random forests would improve concordance versus mean
imputation was 0.482 and 0.962, respectively.
\newline\textbf{Conclusion} Selecting an optimal strategy to impute
missing values and applying multiple imputation can improve the
prognostic accuracy of downstream risk prediction models.
\end{abstract}

\keywords{
    Missing Data
   \and
    INTERMACS
   \and
    imputation
   \and
    heart failure
   \and
    mortality
   \and
    risk prediction
  }

@@

Heart disease is a leading cause of death in the United States. Heart
failure, a primary component of heart disease, affects over 6 million
Americans, and for \textasciitilde10\% of these patients medical
management is no longer effective
\cite{benjamin2017heart,national2017health}. Mechanical circulatory
support (MCS) is a surgical intervention in which a mechanical device is
implanted in parallel to the heart to improve circulation
\cite{patel2014contemporary}. Typically, MCS is used while a patient
waits for a heart transplant (bridge-to-transplant) or in some cases as
an alternative to transplant (destination therapy)
\cite{slaughter2009advanced}. Over 250,000 patients could benefit from
MCS \cite{miller2011left}. However, less than 4,000 new patients receive
a long-term MCS device each year, with widely heterogeneous outcomes
\cite{stewart2011keeping}. The 2-year survival probability on MCS ranges
from 61\% for destination therapy to 78\% for bridge-to-transplant
\cite{patel2014contemporary}. Reliable predictions of patient-specific
risk to experience key outcomes such as mortality or transplant after
receiving MCS can help improve patient selection, inform the design of
next generation pumps, and improve patient care strategies.

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
A critical step for developing risk prediction models is handling
missing data, which is common in a ``real world'' database that is not
designed for the rigor of data completeness found in a clinical trial.
While some missing values are related to data entry error, others may
relate to instability of a patient's circulatory status at the time of
implant. For instance, hemodynamic laboratory values are missing for
patients in whom invasive catheterizations were not performed. Patients
without invasive catheterizations may be too ill (unstable) to tolerate
the procedure. Instead, such patients may, after determination of basic
hemodynamics, proceed directly to device placement.

@@

@@

INTERMACS is a North American observational registry for patients
receiving MCS devices that began as a partnership between the National
Heart, Lung, and Blood Institute, US Food and Drug Administration, the
Centers for Medicaid and Medicare Services, industry, and individual
hospitals with the mission of improving MCS outcomes. In 2018, INTERMACS
became an official Society of Thoracic Surgeons database.

The current analysis was conducted using publicly available data
provided by the National Heart, Lung, and Blood Institute (see
\url{https://biolincc.nhlbi.nih.gov/studies/intermacs/}). We included a
contemporary cohort of 14,738 patients who received continuous flow LVAD
from 2012-2017. This is a secondary analysis of de-identified data
obtained from the National Heart Lung and Blood Institute. Primary data
collection is approved through University of Alabama Institutional
Review Board and at individual sites.

\hypertarget{outcomes-and-predictors}{%
\subsection{Outcomes and Predictors}\label{outcomes-and-predictors}}

Patient follow-up begins after implantation of a durable, long term MCS
device and continues while the device is in place. Registry endpoints
include death on a device, heart transplantation, or cessation of
support (for recovery and non-recovery reasons). Mortality and
transplant after MCS were the primary outcomes for the current study. As
there were only 310 cessation of support events, we did not analyze this
outcome. For mortality and transplant risk prediction models, we applied
cause-specific analysis to account for competing risks. For example, in
risk prediction models for mortality, patients were censored at time of
loss to follow-up, transplant, or cessation of support, whichever
occurred first. INTERMACS collects pre-implant data on patient
characteristics, medical status, and laboratory values. INTERMACS also
collects follow-up data at regularly scheduled visits and at occurrence
of adverse events such as re-hospitalization. For the current analysis,
all pre-implant variables were considered as potential predictors.

@@

To conduct statistical inference in the context of missing data setting,
analysts often create multiple imputed datasets, replicate an analysis
in each of them, and then pool their results to obtain valid test
statistics for hypothesis testing \cite{rubin2004multiple}. Imputation
strategies that create a single imputed dataset
(\textit{e.g., }imputation to the mean) have been shown to increase type
I errors (\textit{i.e., }rejecting a true null hypothesis) for
inferential statistics by artificially reducing the variance of observed
data and ignoring the uncertainty attributed to missing values
\cite{van2018flexible}. To ensure valid inference, imputation models
should leverage outcome variables to predict missing predictors
\cite{sterne2009multiple}. When very few data are missing, analysts may
apply listwise deletion, \textit{i.e., }removing any observation with at
least one missing value. However, listwise deletion can easily lead to
biased inference \cite{van2020rebutting}.

@@

Several of the imputation methods we considered allowed for creation of
single or multiple imputed datasets. For downstream models fitted to
multiple sets of imputed data, we applied the modeling technique to each
imputed training set, separately, and then used a pooling technique to
generate a single set of predictions based on multiple imputed testing
data. Specifically, we created 10 imputed training and testing sets,
then applied a modeling procedure to each imputed training set and
computed model predictions on the corresponding imputed testing set,
which led to 10 sets of predictions. To aggregate these predictions, we
computed the median for each patient. Informal experiments where the
mean was used instead of the median to aggregate predictions showed
little or no difference in the model's prediction accuracy.

@@

Imputing data to the mean involves three steps. First, numeric and
nominal variables are identified. Second, for each numeric variable, the
mean is computed and used to impute missing values in the corresponding
variable. Third, for each nominal variable, the mode is computed
(because one cannot compute a mean for categorical variables) and used
to impute missing values in the corresponding variable. The computed
means and modes are then stored for future imputation of testing data.
Because imputation to the mean is frequently used in practice, we use it
as a reference for comparison of all other strategies to impute missing
values.

@@

Imputation with Bayesian regression draws imputed values from the
posterior distribution of model parameters, accounting for uncertainty
in both model error and estimation \cite{rubin2004multiple}. Multiple
imputation with chained equations (MICE), a well known technique for
generating multiply imputed data
\cite{azur2011multiple, van2018flexible, van2006fully}. In each
iteration of MICE, each specified variable in the dataset is imputed
using other variables in the dataset. These iterations are run until
convergence criteria have been met.

@@

\hypertarget{evaluating-imputation-accuracy}{%
\subsection{Evaluating Imputation
Accuracy}\label{evaluating-imputation-accuracy}}

@@

Imputation accuracy was computed for each numerical and nominal
variable, separately. Numeric variable imputation accuracy was measured
using a re-scaled mean-squared error:
\[1 - \frac{\textrm{MSE}(\textrm{current imputation method})}{\textrm{MSE}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is smaller than
\(\textrm{MSE}(\textrm{imputation to the mean})\), equal to 0 if the two
MSEs are equal, and less than 0 if
\(\textrm{MSE}(\textrm{current imputation method})\) is greater than
\(\textrm{MSE}(\textrm{imputation to the mean})\). Nominal variable
imputation accuracy was measured using a re-scaled classification
accuracy:
\[1 - \frac{\textrm{Classification error}(\textrm{current imputation method})}{\textrm{Classification error}(\textrm{imputation to the mean})}.\]
This score is greater than 0 if the classification error of the current
imputation method is less than that of imputation to the mean, equal to
0 if the two classification errors are equal, and less than 0 of the
classification error of the current imputation method is worse than
imputation to the mean. These numeric and nominal scores are analogous
to the more well known \(R^2\) and Kappa statistics, respectively, but
are modified slightly in the current analysis so that imputation to the
mean will always have a score of 0. This modification makes it easier to
compare the accuracy of each imputation strategy directly with that of
imputation to the mean.

@@

\hypertarget{computational-details}{%
\subsection{Computational Details}\label{computational-details}}

@@

SAS software (version 9.4) and Python (version 3.8.2) were used to
create analytic data for the current study
\cite{van1995python, sas2013}. Base R (version 4.0.3) was used in
combination with a number of open-source R packages
(\textit{e.g., }\texttt{drake}, \texttt{tidyverse}, \texttt{naniar},
\texttt{table.glue}, \texttt{mice}, \texttt{miceRanger}, and others) to
conduct our analysis
\cite{drake, naniar, mice, miceRanger, table.glue, tidyverse, rstanarm, tidybayes, survival, tidymodels, riskRegression}.
Our R code is available on GitHub (see
\url{https://github.com/bcjaeger/INTERMACS-missing-data})
\cite{byron_2020_4247449}. We used Cheaha, a high performance computing
cluster at University of Alabama at Birmingham, to perform the 200
Monte-Carlo cross validation runs \cite{cheaha}.

@@

\hypertarget{bayesian-analysis-of-model-performance}{%
\subsection{Bayesian analysis of model
performance}\label{bayesian-analysis-of-model-performance}}

Adjusting for the amount of additional missing data amputed and the
outcome variable, the posterior probability that an imputation strategy
would improve the scaled Brier score of a downstream model relative to
imputation to the mean was maximized by using multiple imputation with
random forests (probability of improvement: 1.000; Figure
\ref{fig:fig_md_strat_infer_ipa}). Similarly, multiple imputation using
random forests was estimated to have the highest posterior probability
of improving the C-index in comparison to using imputation to the mean
(probability of improvement: 1.000; Figure
\ref{fig:fig_md_strat_infer_auc}). However, the estimated posterior
probability of a reduction in calibration was 1 when using either single
or multiple imputation with random forests compared with imputation to
the mean (Figure \ref{fig:fig_md_strat_infer_cal_error}). Although
imputation using random forest was estimated to be the best overall
option, there was moderate to strong evidence that MIA and each multiple
imputation strategy we applied would improve prognostic accuracy of
downstream models compared with imputation to the mean.

@@

In this article, we leveraged INTERMACS registry data to evaluate how
the use of different imputation strategies prior to fitting a risk
prediction model would impact the external prognostic accuracy of the
model. External prognostic accuracy was measured at 6 months after
receiving MCS, and the primary measure of accuracy was the scaled Brier
score. We evaluated the performance of 12 imputation strategies in a
broad range of settings by varying (1) the amount of additional missing
data amputed prior to performing imputation, (2) the type of risk
prediction model applied after imputation, and (3) the outcome variable
for the risk prediction model. Our resampling experiment indicated that
conducting multiple imputation has a high likelihood of increasing the
downstream scaled Brier score and C-index of risk prediction models
compared with imputation to the mean. Additionally, multiple imputation
with random forests emerged as the imputation strategy that maximized
the probability of developing a more prognostic model compared with
imputation to the mean.

In previous studies involving the INTERMACS data registry, imputation to
the mean has been applied prior to developing a mortality risk
prediction model
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
An interesting recent study indicates that imputation to the mean can
provide an asymptotically consistent prediction model, given the
prediction model is flexible and non-linear \cite{josse2019consistency}.
However, theoretical results for finite samples have not yet been
established. Our results provide relevant data for the finite sample
case, suggesting that using imputation strategies considered in the
current study instead of imputation to the mean can improve the
prognostic accuracy of downstream models, particularly if multiple
imputation is applied. Imputation to the mean should be avoided in
future analyses of the INTERMACS registry and analyses where inflexible
models are applied.

Previous research has also established evidence in favor of applying
multiple imputation to improve the prognostic value of risk prediction
models. For example, Hassan and Atiya demonstrated superior downstream
prediction using an ensemble multiple imputation method on synthetic
data with continuous outcomes \cite{hassan2007regression}. Similarly,
Nanni et. al demonstrated superior performance in downstream prediction
when missing values were imputed using their proposed ensemble multiple
imputation method \cite{nanni2012classifier}. Notably, the authors
artificially induced missing values in these studies and the largest
real dataset that was evaluated contained less than 700 observations. An
article by Jerez et. al evaluated missing data strategies based on the
downstream task of fitting a neural network and predicting early breast
cancer relapse \cite{jerez2010missing}. The authors found that KNN
imputation led to risk prediction models with the highest discrimination
and lowest calibration error. Results from the current study are
consistent with these previous findings but also extend their results by
providing evidence from a larger source of data
(\textit{i.e., }INTERMACS) and dealing with `real-world' missing values.

Others have previously evaluated imputation techniques based on the
accuracy with which these techniques impute missing values in the
training data
\cite{tutz2015improved, little2013joys, steele2018machine}. While it is
intuitive to hypothesize that more accurate imputation will provide more
prognostic downstream models, our results do not support this
supposition. For example, when an additional 30\% of missing data were
amputed, none of the missing data strategies we implemented obtained
higher accuracy than imputation to the mean. However, using \emph{any}
of the multiple imputation strategies we considered instead of
imputation to the mean increased prognostic accuracy of downstream
models when an additional 30\% of missing data were amputed. This result
is likely explained by the bias variance tradeoff. In particular, single
imputation techniques may lead to prediction models with lower bias but
higher variance than multiple imputation techniques.

\paragraph{Strengths and limitations}

The current analysis has a number of strengths. We leveraged the
INTERMACS data registry, comprising one of the largest cohorts of
patients who received MCS. We applied a well known resampling method to
internally validate modeling algorithms for risk prediction. Last, we
made our analysis R code available in a public repository
(\url{https://github.com/bcjaeger/INTERMACS-missing-data}). Last, the
approach presented in this paper provides a general framework that can
be applied in other studies where missing data are imputed prior to
fitting a risk prediction model. The current analysis should also be
interpreted in the context of known limitations. We considered a small
subset of existing strategies to impute missing data, and other
strategies may have provided stronger improvements compared with
imputation to the mean. The models we studied obtained low values of
scaled Brier score, indicating low prediction accuracy. It is unclear
how results may vary for models with higher prediction accuracy. Also,
it was not feasible to use only the training data to impute missing
values in the testing data due to a lack of available software. Although
the \texttt{miceRanger} package allows imputation of new data using
existing models, few software packages for imputation allow users to
implement multiple imputation with this protocol. Future analyses should
introduce more flexible software and hands-on tutorials so that future
investigators can optimize imputation of missing data.

\paragraph{Conclusion}

Selecting an optimal strategy to impute missing values can impact the
prognostic accuracy of downstream risk prediction models. In the current
analysis, conducting multiple imputation using random forests emerged as
an optimal strategy to impute missing values in the INTERMACS data. This
investigation can directly inform future analyses of INTERMACS data and
provide evidence quantifying the benefit of imputing missing data with
sound methodology.

\clearpage
\begin{table}
\caption{Participant characteristics stratified by event status. The majority of patients were male and nearly all mechanical circulatory support devices were left-ventricular assistance devices.}
\label{tbl_characteristics}

\begin{tabular}{lccccc}
\toprule
Characteristic & Overall & Censored & Cessation of support & Dead & Transplant\\
\midrule
No. of patients & 14,738 & 6,308 & 310 & 4,204 & 3,916\\
Age, years & 57 (13) & 57 (13) & 48 (14) & 61 (12) & 53 (12)\\
Sex &  &  &  &  & \\
\hspace{1em}Female & 21 & 22 & 41 & 22 & 19\\
\hspace{1em}Male & 79 & 78 & 59 & 78 & 81\\
\addlinespace
Race &  &  &  &  & \\
\hspace{1em}White & 66 & 63 & 64 & 71 & 67\\
\hspace{1em}Black & 25 & 28 & 27 & 21 & 23\\
\hspace{1em}Other & 9.2 & 9.6 & 9.4 & 7.4 & 10\\
Body mass index & 29 (7) & 29 (7) & 29 (8) & 29 (7) & 28 (6)\\
\addlinespace
Device strategy &  &  &  &  & \\
\hspace{1em}Other & 0.6 & 0.6 & 2.6 & 0.5 & 0.6\\
\hspace{1em}Bridge to transplant & 52 & 43 & 51 & 40 & 81\\
\hspace{1em}Destination therapy & 47 & 56 & 46 & 60 & 18\\
LVEDD & 6.83 (1.10) & 6.85 (1.07) & 6.49 (0.93) & 6.69 (1.12) & 6.95 (1.12)\\
\addlinespace
Urinary albumin, g/dl & 3.41 (0.64) & 3.42 (0.61) & 3.41 (0.64) & 3.33 (0.66) & 3.49 (0.65)\\
Urinary creatinine, mg/dl & 1.39 (0.68) & 1.36 (0.62) & 1.17 (0.53) & 1.49 (0.77) & 1.35 (0.67)\\
CV pressure & 10.8 (6.2) & 10.4 (6.1) & 9.9 (6.0) & 11.6 (6.5) & 10.7 (6.0)\\
Periphal edema & 35 & 33 & 37 & 44 & 31\\
BUN, mg/dl & 29 (18) & 28 (17) & 22 (13) & 32 (19) & 28 (18)\\
\addlinespace
Bilirubin levels, mg/dl & 1.37 (1.84) & 1.25 (1.14) & 1.28 (1.95) & 1.52 (2.46) & 1.41 (1.97)\\
Device type &  &  &  &  & \\
\hspace{1em}Bi-ventricular assistance device & 3.5 & 2.1 & 0.3 & 5.8 & 3.4\\
\hspace{1em}Left-ventricular assistance device & 97 & 98 & 100 & 94 & 97\\
Surgery time, minutes & 292 (113) & 289 (110) & 272 (96) & 320 (128) & 272 (98)\\
\addlinespace
CPB time, minutes & 95 (49) & 93 (45) & 88 (38) & 103 (56) & 90 (44)\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Table values are mean (standard deviation) or \%}\\
\end{tabular}
